Optical characterization of gaps in directly bonded Si compound optics using infrared spectroscopy

### 2015-01-05 AO_bondedSi.tex
Incorporate DTJ edits on pages 1-8

\begin{abstract}
Silicon direct bonding offers flexibility in the design and development of Si optics by allowing manufacturers to combine subcomponents with a potentially lossless and mechanically stable interface. The bonding process presents challenges in meeting the requirements for optical performance because air gaps at the Si interface cause large Fresnel reflections. Even small (35 nm) gaps reduce transmission through a direct bonded Si compound optic by 5\% at $\lambda = 1.15 \; \mu$m at normal incidence. Understanding the optical effects of such gaps and evaluation of various methods for preventing or eliminating them demands that we have a method for measuring not only the two-dimensional extent of the gaps but also their axial extent.  Existing methods for studying gaps provide only poor measures of this axial extent. We can use such instruments to detect and measure gaps as small as 8 nm by measuring transmission as a function of wavelength. We describe a bond inspection method that makes use of precision transmission spectrophotometry. Our analysis of the spectra then produces accurate estimates of the axial depth of air gaps.  Our approach involves modeling multiple incoherent reflections within the substrate interfaces with a wave transfer matrix, modified for intensities rather than complex amplitudes. We describe the measurement and analysis techniques and demonstrate the validity of the approach by measuring bond gaps of known depths produced by microlithography.
\end{abstract}

@@

\subsection{Why bonding is good and AR Coating is bad}
As with a number of glasses and crystalline optical materials, it is possible to form strong physical bonds between two planar or conformal silicon surfaces.  One advantage of this direct Si-Si bonding of macroscopic optical parts is that it facilitates the manufacture of complex optical subsystems, for example double-convex aspheres, combinations of lenses, prisms, and transmission gratings, and pairs of gratings oriented at right angles \verb|\cite{2012SPIE.8450E..2TV, 2010SPIE.7739E.123G}|.  A second important advantage is that a sufficiently close bond is optically lossless, an important consideration not only for its effect on throughput but also because it can eliminate concerns about optical ghosts.  The remaining advantages of bonding stem from various difficulties encountered in antireflection coating a high index material in the near and mid-infrared:  There is a limited choice of index-matching materials, in particular at longer wavelengths.  Systems often are used at cryogenic temperatures and considerations of differential thermal expansion, brittleness and hygroscopic properties can restrict the choices even further. It is hard to get good antireflection performance across broad spectral ranges and at a range of incidence angles.

@@

If Si-Si bonds are to be a useful part of the optical designer's toolbox, the bonds must have transmission losses similar to or lower than those of two antireflection coated surfaces over the operating wavelength (typically $< 2$\%).  If the optical layout does not eliminate pupil ghosts from the flat Si surfaces, there are additional requirements on the reflectivity of individual bubbles in the bond. While completely bonded Si parts will be completely lossless at the interface, problems on different scales with different physical origins can lead to less-than-perfect bonds.  On the smallest scales, imperfect cleaning can leave small voids where particulates hold the bonding planes apart or where hydrocarbons can serve as catalysts for gap formation.  Even in the best of cleanrooms, particles at the optically significant 10-20 nm scale can be present in relatively high abundance. Figure 5 of Cooper 1986\verb|\cite{doi:10.1080/02786828608959094}| shows the surface flux (particles cm$^2$ s$^{-1}$) vs. particle size ($\mu$m). There will be roughly 104 times more 10 nm sized particles than 1 $\mu$m sized particles.  On larger scale, gaseous hydrogen, which is a byproduct of the hydrophilic bonding process, can migrate into the bond and form local bubbles \verb|\cite{Masteika2014}|.  Most of the bonding literature concerns itself with wafer bonding.  In bonding wafers to each other or to thicker optically polished Si disks, one or both of the partners can conform as the bonding front propagates.  When bonding two thicker and more rigid substrates, differences in the shapes of the polished surfaces can potentially lead to topological inconsistencies that prohibit some areas from bonding.

\subsection{Existing detection schemes}
Verification that an optical bond meets its requirements and possible diagnosis of problems in the bonding technique require an accurate assessment of the geometry and lateral and axial dimensions of any defects.  Since silicon is opaque in the optical, the most straightforward technique for detecting defects in the bond is to image it in the infrared.  Fresnel losses will cause unbonded regions to appear darker than regions where the bond is perfect and the transmission is 100\%.  In cases where the axial extent of the gap is a wavelength or more, Newton rings appear in the image of the gap.  X-ray topography \verb|\cite{Mitani1990}| exploits the change in optical density along lines of sight to allow precise high spatial resolution measurements of the spatial extent of gaps with axial dimensions as small as a few nanometers but not the axial extent itself. In ultrasound microscopy, the density change at the gap interface produces reflections of sound waves and enables mapping the lateral gap dimensions but produces limited information about the axial dimension \verb|\cite{2000RScI...71.1869G}|.

\subsection{Technique}
We describe here a new technique for measuring the axial extent of gaps, founded upon the ability to measure transmission as a function of wavelength with high accuracy.   Stock near-infrared spectrophotometer like the Cary5000 from Agilent offer 0.2\% precision and good repeatability.  We use the wavelength dependence of the transmission and a model of the gap as a closely-spaced Fabry-Perot to measure axial gaps down to a few nanometers in size.  In order to verify the applicability and accuracy of the technique, we have used microlithography to create small artificial gaps of known size in wafers that we then bond and measure.  Nowhere in the literature did we find a technique similar to the one presented here, probably because the technique relies on the high precision of relatively new IR capable spectroscopy that was not widely available until this decade.

\subsection{What level do you need to measure to}
For all types of gaps and bubbles, our technique needs to provide an accurate assessment of the transmission loss at the operating wavelength.  For finite-sized bubbles, measurement of the axial dimension helps us to assess the effects of high-temperature annealing on removal of hydrogen bubbles.  A wavelength-dependent transmission variation can also indicate the presence of large numbers of gaps that are too small to resolve spatially, as one might expect if there are many very small contaminants present along the interface.  To meet these diagnostic goals, our technique must be able to characterize gaps with losses of a few percent at the operating wavelength.  A 2\% loss at 1.15 $\mu$m corresponds to a gap of XX nm at normal incidence and the gap size for a loss of this magnitude scales linearly with wavelength.  We therefore need to be able to measure transmission as a function of wavelength to a fraction of a percent, including all systematic errors, for devices operating near 1 $\mu$m and to a somewhat looser standard for devices operating at longer wavelengths.

@@

Another possible source of gaps at the bond is inherent roughness in the initial surfaces that leads to a failure to conform.  Stiff non-flat Si surfaces will not conform upon contact leaving the complement of the aberrations as an air gap. Unconventional thick ($\sim$ 30 mm) Si pucks are less conformal than wafers are, so we expect thick substrates could have gaps even in the absence of particles.  Most of our thick substrates are optically polished to less than 60 nm peak to valley over the 3 to 4 inch diameter clear aperture.  We assume that if enough pressure is placed on the substrates during bonding, then the surfaces will come in contact and zipper up.  Subsequent annealing will strength the bond.

@@

\begin{figure}[htbp]
\centerline{\includegraphics[width=0.95\columnwidth]{figs/SiIndexAOmgsFinesseFig.pdf}}
\caption{Plot of refractive index, $n$, and coefficient of finesse, $F$, as a function of wavelength, $\lambda$.\label{figSiIndexFinesse} We computed the temperature dependent refractive index from the tabulations of CITE\texttt{2006SPIE.6273E..77F}.  The coefficient of finesse is defined in Equation \ref{eq:FabPerot}.  The minuscule temperature dependence is largely negligible, and for the rest of the paper we assume room temperature refractive index.}
\end{figure}

The Si$-$Si interface gap is a low-Finesse Fabry-P\`{e}rot etalon\verb|\cite{2007fuph.book.....S}|.  The transmission through a Fabry-P\`{e}rot depends on the wavelength of light, the reflectivity of the etalon sidewalls, and the size of the gap.  We assume transmission is at normal incidence, the refractive index of the gap is 1.0, and the material is at room temperature.  We compute the reflectivity for an Si$-$vacuum interface, which is a modest function of wavelength through the refractive index of Silicon, $n(\lambda, T)$ \verb|\cite{2006SPIE.6273E..77F}|, as shown in Figure \ref{figSiIndexFinesse}.  In the equations ahead we drop the subscripts on the refractive index $n(\lambda, T) \rightarrow n$ and the coefficient of finesse $F(\lambda, T) \rightarrow F$ for clarity.  The computation is simply Fresnel's law at normal incidence:

>>>>

\begin{abstract}
Silicon direct bonding offers flexibility in the design and development of Si optics by allowing manufacturers to combine subcomponents with a potentially lossless and mechanically stable interface. The bonding process presents challenges in meeting the requirements for optical performance because air gaps at the Si interface cause large Fresnel reflections. Even small (35 nm) gaps reduce transmission through a direct bonded Si compound optic by 5\% at $\lambda = 1.15 \; \mu$m at normal incidence. Understanding the optical effects of such gaps and evaluation of various methods for preventing or eliminating them demands that we have a method for measuring not only the two-dimensional extent of the gaps but also their axial extent.  Existing methods for studying gaps provide only poor measures of this axial extent.  We describe a bond inspection method that makes use of precision transmission spectroscopy.  Using spectroscopy, we can detect and measure gaps as small as 8 nm by measuring transmission as a function of wavelength. Our approach involves modeling multiple incoherent reflections within the substrate interfaces with a wave transfer matrix, modified for intensities rather than complex amplitudes. We describe the measurement and analysis techniques and demonstrate the validity of the approach by measuring bond gaps of known depths produced by microlithography.
\end{abstract}

@@

\subsection{Why bonding is good and AR Coating is bad}
As with a number of glasses and crystalline optical materials, it is possible to form strong physical bonds between two planar or conformal silicon surfaces.  One advantage of this direct Si-Si bonding of macroscopic optical parts is that it facilitates the manufacture of complex optical subsystems, for example double-convex aspheres, combinations of lenses, prisms, and transmission gratings, and pairs of gratings oriented at right angles \verb|\cite{2012SPIE.8450E..2TV, 2010SPIE.7739E.123G}|.  A second important advantage is that a sufficiently close bond is optically lossless, an important consideration not only for its effect on throughput but also because it can eliminate concerns about optical ghosts.  The remaining advantages of bonding stem from various difficulties encountered in antireflection coating a high index material in the near and mid-infrared:  There is a limited choice of index-matching materials, in particular at longer wavelengths.  Systems often are used at cryogenic temperatures and considerations of differential thermal expansion, brittleness and hygroscopic properties can restrict the choice of coatings even further. It is hard to get good antireflection performance across broad spectral ranges and at a range of incidence angles.

@@

If Si-Si bonds are to be a useful part of the optical designer's toolbox, the bonds must have transmission losses similar to or lower than those of two antireflection coated surfaces over the operating wavelength (typically $< 2$\%).  If the optical layout does not eliminate pupil ghosts from the flat Si surfaces, there are additional requirements on the reflectivity of individual bubbles in the bond. While completely bonded Si parts will be completely lossless at the interface, problems on different scales with different physical origins can lead to less-than-perfect bonds.  On the smallest scales, imperfect cleaning can leave small voids where particulates hold the bonding planes apart or where hydrocarbons can serve as catalysts for gap formation.  Even in the best of cleanrooms, particles at the optically significant 10-20 nm scale can be present in relatively high abundance. Figure 5 of Cooper 1986\verb|\cite{doi:10.1080/02786828608959094}| shows the surface flux (particles cm$^2$ s$^{-1}$) vs. particle size ($\mu$m). There will be roughly $10^4$ times more 10 nm sized particles than 1 $\mu$m sized particles.  On larger scale, gaseous hydrogen, which is a byproduct of the hydrophilic bonding process, can migrate into the bond and form local bubbles \verb|\cite{Masteika2014}|.  Most of the bonding literature concerns itself with wafer bonding.  In bonding wafers to each other or to thicker optically polished Si disks, one or both of the partners can conform as the bonding front propagates.  When bonding two thicker and more rigid substrates, differences in the shapes of the polished surfaces can potentially lead to topological inconsistencies that prohibit some areas from bonding.

\subsection{Existing detection schemes}
Verification that an optical bond meets its requirements and diagnosis of possible problems in the bonding technique require an accurate assessment of the geometry and lateral and axial dimensions of any defects.  Since silicon is opaque in the optical, the most straightforward technique for detecting defects in the bond is to image it in the infrared.  Fresnel losses will cause unbonded regions to appear darker than regions where the bond is perfect and the transmission is 100\%.  In cases where the axial extent of the gap is a wavelength or more, Newton rings appear in the image of the gap.  X-ray topography \verb|\cite{Mitani1990}| exploits the change in optical density along lines of sight to allow precise high spatial resolution measurements of the spatial extent of gaps with axial dimensions as small as a few nanometers but not the axial extent itself. In ultrasound microscopy, the density change at the gap interface produces reflections of sound waves and enables mapping the lateral gap dimensions but produces limited information about the axial dimension \verb|\cite{2000RScI...71.1869G}|.

\subsection{Technique}
We describe here a new technique for measuring the axial extent of small gaps, founded upon the ability to measure transmission as a function of wavelength with high accuracy.   Stock near-infrared spectrophotometers like the Cary5000 from Agilent offer 0.2\% precision and good repeatability.  We use the wavelength dependence of the transmission and a model of the gap as a closely-spaced Fabry-Perot to measure axial gaps down to a few nanometers in size.  In order to verify the applicability and accuracy of the technique, we have used microlithography to create small artificial gaps of known size in wafers that we then bond and measure.  Nowhere in the literature did we find a technique similar to the one presented here, probably because the technique relies on the high precision of relatively new IR capable spectroscopy that was not widely available until this decade.

\subsection{What level do you need to measure to}
For all types of gaps and bubbles, our technique needs to provide an accurate assessment of the transmission loss at the operating wavelength.  For finite-sized bubbles, measurement of the axial dimension helps us to assess the effects of high-temperature annealing on removal of hydrogen bubbles.  A wavelength-dependent transmission variation can also indicate the presence of large numbers of gaps that are too small to resolve spatially, as one might expect if there are many very small contaminants present along the interface.  An important figure of merit for a bond is whether any loss due to gaps is smaller than the loss one might expect from two passages through a reasonably effective broad-band anti-relfection coating in the infrared.  To meet this diagnostic goal, our technique must be able to characterize gaps with losses of a few percent at the operating wavelength.  A 2\% loss at 1.15 $\mu$m corresponds to a gap of XX nm at normal incidence and the gap size for a loss of this magnitude scales linearly with wavelength.  We therefore need to be able to measure transmission as a function of wavelength to a fraction of a percent, including all systematic errors, for devices operating near 1 $\mu$m and to a somewhat looser standard for devices operating at longer wavelengths.

@@

Another possible source of gaps at the bond is inherent roughness in the initial surfaces that leads to a failure to conform.  Stiff non-flat Si surfaces will not conform upon contact leaving the complement of the aberrations as an air gap. Unconventional thick ($\sim$ 30 mm) Si pucks are less conformal than wafers are, so we expect thick substrates could have gaps even in the absence of particles.  Most of our thick substrates are optically polished to less than 60 nm peak to valley over the 3 to 4 inch diameter clear aperture.  We assume that, if enough pressure is placed on the substrates during bonding, then the surfaces will come in contact and zipper up.  Subsequent annealing will strength the bond.

@@

\begin{figure}[htbp]
\centerline{\includegraphics[width=0.95\columnwidth]{figs/SiIndexAOmgsFinesseFig.pdf}}
\caption{Plot of refractive index, $n$, and coefficient of finesse, $F$, as a function of wavelength, $\lambda$.\label{figSiIndexFinesse} We computed the temperature dependent refractive index from the tabulations of CITE\texttt{2006SPIE.6273E..77F}.  The coefficient of finesse is defined in Equation \ref{eq:FabPerot}.  We assume $T= 295\;$K for all calculations in this article.}
\end{figure}

The Si$-$Si interface gap is a low-Finesse Fabry-P\`{e}rot etalon\verb|\cite{2007fuph.book.....S}|.  The key property of an etalon is that it has a cavity enclosed between two smooth and parallel reflective surfaces where the requirements for smoothness and parallelism depend on the reflective finesse of the cavity.  For the narrow air gaps we consider here and for the modest reflectivity of a Si-vacuum interface, most gaps at Si-Si bonds will form effective etalons over most of their area.  The transmission through a Fabry-P\`{e}rot depends on the wavelength of light, the reflectivity of the etalon sidewalls, and the size of the gap.  We assume transmission is at normal incidence, the refractive index of the gap is 1.0, and the material is at $T= 295\;$K.  We compute the reflectivity for an Si$-$vacuum interface, which is a modest function of wavelength through the refractive index of Silicon, $n(\lambda, T)$ \verb|\cite{2006SPIE.6273E..77F}|, as shown in Figure \ref{figSiIndexFinesse}.  In the equations ahead we drop the subscripts on the refractive index $n(\lambda, T) \rightarrow n$ and the coefficient of finesse $F(\lambda, T) \rightarrow F$ for clarity.  The computation is simply Fresnel's law at normal incidence:
